---
title: "COVID-19 modeling in South Dakota"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: github_document
---
# Authors
*Jeff Wesner, Ph.D.*^1^, *Dan Van Peursem, Ph.D.*^2^, *Jose Flores, Ph.D.*^2,3^, *Yuhlong Lio, Ph.D.*^2^

University of South Dakota

^1^Department of Biology, ^2^Department of Mathematical Sciences, ^3^Department of Computer Science

<Jeff.Wesner@usd.edu>

# Purpose
To predict hospital bed needs in South Dakota due to COVID-19. 

# Updates

# Alternative Approach
## Directly modeling hospital bed use

Our previous approach attempted to estimate hospital bed use from an initial SIR (https://github.com/jswesner/covid_sd/blob/master/code/alternative_analysis.pdf). However, there is now enough data to directly model the curve of cumulative COVID related hospitalizations. To do that, we first used the Weibull equation to fit a Bayesian non-linear curve to cumulative hospitalization data (first graph below). We then converted that cumulative prediction to active hospitalizations by varying the days in the hospital to fit the data on active hospitalizations.   
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(viridis)
library(deSolve)
library(scales)
library(lubridate)
library(viridis)
library(readr)
library(readxl)
library(ggrepel)
library(cowplot)
library(viridis)
library(brms)
library(kableExtra)
```


```{r echo=FALSE, fig.height=2, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot() +
  annotate("text", x = 0.5, y = 0.7,
                   label = "(y_cumulative[i]) %~% italic(Poisson)(lambda[i])", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.52,
           label = "italic(log(lambda[i])) == frac(italic(maxX)*date[i]^omega, theta^omega + date[i]^omega)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.35,
           label = "italic(max) %~% italic(N)(8,1)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.27,
           label = "italic(theta) %~% italic(N)(50, 10)", parse = T, family = "serif") +
   annotate("text", x = 0.5, y = 0.19,
           label = "italic(omega) %~% italic(Gamma)(4, 2)", parse = T, family = "serif") +
  coord_cartesian(ylim = c(0.11, .8)) +
  labs(title = "Weibull non-linear regression:") +
  theme_void() +
  theme(text = element_text(family = "serif"))
```

Prior values for are loosely derived from from New York City's hospitalization curve in which ~ 0.4% of the population was hospitalized. We used a poisson likelihood because the data are positive counts and the variance is likely to increase with the mean.

## Weibull results

We converted the posterior prediction of cumulative hospitalizations to a prediction of active hospitalizations (number of people hospitalized on any given day). We did that by assuming four scenarios of average hospital stay lengths (5, 10, 12, and 15 days). Those predictions were plotted against the actual active hospitalizations reported from the South Dakota Deparment of Health. Results from the model suggest that South Dakota either has or shortly will reach its peak in COVID-related hospitalizations at ~100 people actively hospitalized. That prediction is most consistent with an average hospital stay of 10-12 days. 

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
sd_covid_data <- read.csv(here::here("data/data_kelo.csv")) %>%
  clean_names() %>%
  mutate(date_num = as.numeric(mdy(date)) - 18329,
         date = mdy(date),
         source = "SD_DOH via Keloland news",
         incidence = positive_cases - lag(positive_cases),
         cum_hosp = parse_number(as.character(hospitalized)))  ##adds 1 to all incidence to allow for zeros in log-linear regression


# max_prop_hosp <- max(nyc_hosp$cum_prop_hosp)
# sd_prior_max = sd_pop*max_prop_hosp
# sd_prior_maxsd = 2000 

cum_hosp <- read_csv("data/cum_hosp.csv", na = c("", ".")) %>% 
  mutate(date = mdy(date_hosp),
         date_numeric  = as.numeric(date),
         date_num = date_numeric - min(date_numeric)) 

# get_prior(bf(cum_hosp ~ (ult*date_num^omega)/(theta^omega + date_num^omega),
#              ult ~ 1 + (1|group), omega ~ 1 + (1|group), theta ~ 1 + (1|group),
#              nl = TRUE),
#           data = cum_hosp,
#           family = poisson(link = "log"))
# 
# gamma_parms <- function(mean, sd) {
#   out <- tibble(shape = mean^2/sd^2,
#                 rate = mean/sd^2,
#                 scale = 1/(mean/sd^2))
#   return(out)
#   return(plot)
# }
# 
# gamma_parms(mean = 2, sd = 1)

cum_hosp_state <- cum_hosp %>% group_by(date, date_num) %>% summarize(cum_hosp = sum(cum_hosp))

#fit weibull with poisson likelihood
# fit_weib_state <- brm(bf(cum_hosp ~ (ult*date_num^omega)/(theta^omega + date_num^omega),
#                          ult ~ 1 , omega ~ 1 , theta ~ 1 ,
#                          nl = TRUE ),
#                       data = cum_hosp_state,
#                       family = poisson(link = "log"),
#                       prior = c(
#                         prior(normal(8,1), nlpar = "ult"), #asymptote - based on proportion total hosp in NY
#                         prior(normal(50, 10), nlpar = "theta"), #time (days since start) of inflection
#                         prior(gamma(4,2), nlpar = "omega")), # slope at inflection. Gamma to ensure it is positive (has to be since data are cumulative) assumes slope of exp(2) with sd of 0.5
#                       iter = 1000, chains = 1,
#                       sample_prior = "yes")

#re-run model
# fit_weib_state <- update(readRDS(here::here("outputs/fit_weib_state.rds")), newdata = cum_hosp_state)

yesterday <- readRDS(here::here("outputs/fit_weib_state.rds"))
post_yest <- posterior_samples(yesterday)
post_yestsum <- post_yest %>% 
  gather() %>% 
  group_by(key) %>% 
  summarize(mean = mean(value),
            sd = sd(value))

post_yestsum$mean[2]

#fit model again with new data and with yesterday's posterior as the prior
fit_weib_state <- update(readRDS(here::here("outputs/fit_weib_state.rds")), newdata = cum_hosp_state,
                         prior = c(
                           prior_string(paste0("normal(",post_yestsum$mean[2],",", post_yestsum$sd[2],")"), nlpar = "theta"),
                           prior_string(paste0("normal(",post_yestsum$mean[3],",", post_yestsum$sd[3],")"), nlpar = "ult"),
                           prior_string(paste0("gamma(" ,post_yestsum$mean[1]^2/post_yestsum$sd[1]^2, ",",post_yestsum$mean[1]/post_yestsum$sd[1]^2,")"), nlpar = "omega")))


saveRDS(fit_weib_state, file = here::here("outputs/fit_weib_state.rds"))
saveRDS(fit_weib_state, file = here::here(paste0("outputs/fit_weib_state_", Sys.Date(), ".rds")))
```

```{r echo=FALSE, fig.height=9, fig.width=8, message=FALSE, warning=FALSE, paged.print=FALSE}
#predictions up to 200 days after the first hospitalization
newdata = tibble(date_num = 0:200)

predict_cumulative <- predict(fit_weib_state, newdata = newdata, probs = c(0.05, 0.95, 0.25, 0.75))  %>% 
  as_tibble() %>% 
  clean_names() %>% 
  mutate(date_num = 18329 + newdata$date_num,
         date = as_date(date_num),
         method = "predicted") 

#make predictions of daily hospital needs
predict_dailyhosp <- predict(fit_weib_state, newdata = newdata, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_num = 18328 + parse_number(key),
         date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(inc = value - lag(value, 1)) %>% 
  mutate(daily_total_5 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4),
         daily_total_10 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9),
         daily_total_12 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11),
         daily_total_15 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11) + lag(inc, 12) + lag(inc, 13) + lag(inc, 14)) %>% 
  gather(hosp_guess, daily_total, c("daily_total_5":"daily_total_15")) %>% 
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  group_by(iter, date, hosp_guess) %>% 
  filter(!any(daily_total < 0)) %>% 
  group_by(date, hosp_guess) %>% 
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high90 = quantile(daily_total, probs = 0.95),
            high50 = quantile(daily_total, probs = 0.75),
            low90 = quantile(daily_total, probs = 0.05),
            low50 = quantile(daily_total, probs = 0.25)) %>% 
  ungroup() %>% 
  mutate(hosp_guess = case_when(hosp_guess == "daily_total_5" ~ "5 days in hospital",
                                hosp_guess == "daily_total_10" ~ "10 days in hospital",
                                hosp_guess == "daily_total_12" ~ "12 days in hospital",
                                TRUE ~ "15 days in hospital"),
         hosp_guess = fct_relevel(hosp_guess, "5 days in hospital"))


cumulative_plot <- predict_cumulative %>% 
  ggplot(aes(x = date, y = estimate)) +
  geom_line(color = "dodgerblue") +
  geom_ribbon(aes(ymin = q5, ymax = q95), alpha = 0.2, fill = "dodgerblue" ) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, fill = "dodgerblue" ) +
  geom_point(data = cum_hosp_state, aes(x = date, y = cum_hosp),
             shape = 21, fill = "yellow") +
  # facet_wrap(~ group, scales = "free") +
  theme_classic() +
  scale_x_date() +
  # scale_x_date(date_breaks = "months" , date_labels = "%b") +
  # ylim(0, 1500) +
  labs(y = "Cumulative ever hospitalized",
       title = "Weibull model results",
       subtitle = "Dots = data. Shading = 50% and 95% prediction intervals") +
  NULL

active_plot <- predict_dailyhosp %>% 
ggplot() +
  geom_line(aes(x = date, y = median), color = "dodgerblue") +
  geom_ribbon(aes(x = date, ymin = low90, ymax = high90), fill = 'dodgerblue', alpha = 0.2) +
  geom_ribbon(aes(x = date, ymin = low50, ymax = high50), fill = 'dodgerblue', alpha = 0.2) +
  geom_point(data = sd_covid_data, aes(x = date, y = hospitalized_currently),
             shape = 21, fill = "yellow") +
  # scale_x_date(date_breaks = "months" , date_labels = "%b") +
  theme_classic() +
  facet_grid(.~hosp_guess) +
  # geom_hline(yintercept = all_beds, color = "dodgerblue") +
  # annotate("text", x = as.Date("2020-04-10"), y = 2960, label = "All Hospital Beds\nin SD") +
  labs(y = "Active hospitalizations",
       title = "Predicting active hospitalizations from Weibull model",
       subtitle = "Shading = 50% and 90% prediction intervals") +
  # coord_cartesian(ylim = c(0,100)) +
  NULL
# active_plot

both_plot_group <- plot_grid(cumulative_plot, active_plot, ncol = 1, align = "v")

both_plot_group

```

## Divergent trends outside of Minnehaha County
The analysis above makes predictions for the state of South Dakota, but most of the early cases in the state happened in Minnehaha County. As shown below, when we plot the hospitalization trends separately, Minnehaha County appears to have nearly peaked in cumulative hospitalizations while the rest of South Dakota is still in the exponential phase of growth. To capture this dynamic, we re-fit the model above but with separate random intercepts for each group (one for Minnehaha County and one for the rest of South Dakota). 

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# fit_weib_group <- brm(bf(cum_hosp ~ (ult*date_num^omega)/(theta^omega + date_num^omega),
#                          ult ~ 1 + (1|group), omega ~ 1 + (1|group), theta ~ 1 + (1|group),
#                          nl = TRUE ),
#                       data = cum_hosp,
#                       family = poisson(link = "log"),
#                       prior = c(
#                         prior(normal(6,1), nlpar = "ult"), #asymptote - lower to account for splitting data
#                         prior(normal(50, 10), nlpar = "theta"), #time (days since start) of inflection
#                         prior(gamma(4,2), nlpar = "omega"),
#                         prior(cauchy(0, 1), class = "sd", nlpar = "ult"),
#                         prior(cauchy(0, 1), class = "sd", nlpar = "theta"),
#                         prior(cauchy(0, 1), class = "sd", nlpar = "omega")))

# plot(conditional_effects(fit_weib_group))

#update models
fit_weib_group <- update(readRDS(here::here("outputs/fit_web_group.rds")), newdata = cum_hosp)

```

```{r echo=FALSE, fig.height=9, fig.width=8, message=FALSE, warning=FALSE, paged.print=FALSE}

#group fits
newdata <- tibble(date_num = seq(0,200, by = 1),
                  group = c("Minnehaha County"))
newdata_rest <- tibble(date_num = seq(0,200, by = 1),
                  group = c("Rest of South Dakota"))

postpreds_minn <- predict(fit_weib_group, newdata = newdata,probs = c(0.95, 0.05, 0.75, 0.25), re_formula = NULL)  %>% 
  as_tibble() %>% clean_names() %>%  mutate(group = "Minnehaha County",
                                            date_num = newdata$date_num)

postpreds_rest <- predict(fit_weib_group, newdata = newdata_rest, probs = c(0.95, 0.05, 0.75, 0.25), re_formula = NULL) %>% 
  as_tibble() %>% clean_names() %>%  mutate(group = "Rest of South Dakota",
                                            date_num = newdata_rest$date_num)


post_all <- bind_rows(postpreds_minn, postpreds_rest)


predict_group_cumulative <- post_all %>% 
  ggplot(aes(x = date_num, y = estimate, fill = group)) +
  geom_line(aes(color = group)) +
  geom_ribbon(aes(ymin = q5,ymax = q95), alpha = 0.2) +
  geom_ribbon(aes(ymin = q25,ymax = q75), alpha = 0.2) +
  geom_point(data = cum_hosp, aes(x = date_num, y = cum_hosp, fill = group), shape = 21) +
  # ylim(0,500) +
  # facet_wrap(~group) +
  # xlim(0, 250) +
  # ylim(0, 1200) + 
  coord_cartesian(ylim = c(0, 500)) + 
  theme_classic() +
  # geom_hline(yintercept = all_beds, color = "dodgerblue") +
  # annotate("text", x = as.Date("2020-04-10"), y = 2960, label = "All Hospital Beds\nin SD") +
  labs(y = "Cumulative hospitalizations",
       title = "Predicting active hospitalizations from Weibull model",
       subtitle = "Shading = 50% and 90% prediction intervals") +
  # coord_cartesian(ylim = c(0,100)) +
  NULL


#active hospitalizations

date_convert <- tibble(date_num = as.numeric(min(cum_hosp$date)) + newdata$date_num) %>% 
  mutate(date_order = 1:nrow(.))

predict_dailyhosp_minn <- predict(fit_weib_group, newdata = newdata, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_order = parse_number(key)) %>% 
  left_join(date_convert) %>% 
  mutate(date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(inc = value - lag(value, 1)) %>% 
  mutate(daily_total_5 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4),
         daily_total_10 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9),
         daily_total_12 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11),
         daily_total_15 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11) + lag(inc, 12) + lag(inc, 13) + lag(inc, 14)) %>% 
  gather(hosp_guess, daily_total, c("daily_total_5":"daily_total_15")) %>% 
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  group_by(iter, date, hosp_guess) %>% 
  filter(!any(daily_total < 0)) %>% 
  group_by(date, hosp_guess) %>% 
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high90 = quantile(daily_total, probs = 0.95),
            high50 = quantile(daily_total, probs = 0.75),
            low90 = quantile(daily_total, probs = 0.05),
            low50 = quantile(daily_total, probs = 0.25)) %>% 
  ungroup() %>% 
  mutate(hosp_guess = case_when(hosp_guess == "daily_total_5" ~ "5 days in hospital",
                                hosp_guess == "daily_total_10" ~ "10 days in hospital",
                                hosp_guess == "daily_total_12" ~ "12 days in hospital",
                                TRUE ~ "15 days in hospital"),
         hosp_guess = fct_relevel(hosp_guess, "5 days in hospital")) %>% 
  mutate(group = "Minnehaha County")



date_convertrest <- tibble(date_num = as.numeric(min(cum_hosp$date)) + newdata_rest$date_num) %>% 
  mutate(date_order = 1:nrow(.))

predict_dailyhosp_rest <- predict(fit_weib_group, newdata = newdata_rest, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_order = parse_number(key)) %>% 
  left_join(date_convertrest) %>% 
  mutate(date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(inc = value - lag(value, 1)) %>% 
  mutate(daily_total_5 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4),
         daily_total_10 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9),
         daily_total_12 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11),
         daily_total_15 = inc + lag(inc, 1) + lag(inc, 2) + lag(inc, 3) + lag(inc, 4) + lag(inc, 5) +
            + lag(inc, 6) + lag(inc, 7) + lag(inc, 8) + lag(inc, 9) + lag(inc, 10) + lag(inc, 11) + lag(inc, 12) + lag(inc, 13) + lag(inc, 14)) %>% 
  gather(hosp_guess, daily_total, c("daily_total_5":"daily_total_15")) %>% 
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  group_by(iter, date, hosp_guess) %>% 
  filter(!any(daily_total < 0)) %>% 
  group_by(date, hosp_guess) %>% 
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high90 = quantile(daily_total, probs = 0.95),
            high50 = quantile(daily_total, probs = 0.75),
            low90 = quantile(daily_total, probs = 0.05),
            low50 = quantile(daily_total, probs = 0.25)) %>% 
  ungroup() %>% 
  mutate(hosp_guess = case_when(hosp_guess == "daily_total_5" ~ "5 days in hospital",
                                hosp_guess == "daily_total_10" ~ "10 days in hospital",
                                hosp_guess == "daily_total_12" ~ "12 days in hospital",
                                TRUE ~ "15 days in hospital"),
         hosp_guess = fct_relevel(hosp_guess, "5 days in hospital")) %>% 
  mutate(group = "Rest of South Dakota")


daily_hosp_all <- bind_rows(predict_dailyhosp_minn, predict_dailyhosp_rest)


active_daily_group <- daily_hosp_all %>% 
  ggplot(aes(x = date, y= median, fill = group)) + 
  geom_line(aes(color = group)) +
  geom_ribbon(aes(fill = group, color = group, ymin = low90, ymax = high90), alpha = 0.2) +
  geom_ribbon(aes(fill = group, color = group, ymin = low50, ymax = high50), alpha = 0.2) +
  # geom_point(data = daily_hosp, aes(x = date, y = daily_hosp, group = group, fill = group), shape = 21) +
  facet_grid(group ~ hosp_guess) +
  theme_classic() +
  # geom_hline(yintercept = all_beds, color = "dodgerblue") +
  # annotate("text", x = as.Date("2020-04-10"), y = 2960, label = "All Hospital Beds\nin SD") +
  labs(y = "Active hospitalizations",
       title = "Predicting active hospitalizations from Weibull model",
       subtitle = "Shading = 50% and 90% prediction intervals") +
  coord_cartesian(ylim = c(0,500)) +
  NULL

both_group <- plot_grid(predict_group_cumulative, active_daily_group, ncol = 1, align = "v")

both_group
```



## Influence of the priors
The plot below compares the prior predictive distribution (i.e. running the model with only the prior information, not the data) to the posterior predictive distribution. The large difference in these predictions indicates that we learned a lot of information from the data and the priors have relatively small influences.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#prior v post

posts <- posterior_samples(fit_weib_state) %>% as_tibble() %>% mutate(iter = 1:nrow(.))

x = seq(1,200, by = 10)

priors <- tibble(prior_b_ult = rnorm(1000, 8, 1),
                 prior_b_omega = rgamma(1000, 4,2), 
                 prior_b_theta = rnorm(1000, 50, 10), 
                 iter = 1:1000)  %>% 
  expand_grid(x) %>% 
  mutate(pred = (prior_b_ult*x^prior_b_omega)/(prior_b_theta^prior_b_omega + x^prior_b_omega),
         model = 'prior_prediction') %>% 
  select(x, pred, model)

posteriors <- posts %>% select(b_ult_Intercept, b_omega_Intercept, b_theta_Intercept)  %>% 
  expand_grid(x) %>% 
  mutate(pred = (b_ult_Intercept*x^b_omega_Intercept)/(b_theta_Intercept^b_omega_Intercept + x^b_omega_Intercept),
         model = 'posterior_prediction') %>% 
  select(x, pred, model)


all_posts <- bind_rows(priors, posteriors) %>% 
  as_tibble() %>% 
  mutate(date_num = 18329 + x,
         date = as_date(date_num))

all_posts %>% 
  group_by(date, model) %>% 
  summarize(mean = mean(exp(pred)),
            low95 = quantile(exp(pred), probs = 0.025),
            high95 = quantile(exp(pred), probs = 0.975)) %>% 
  ggplot(aes(x = date, y = mean, ymin = low95, ymax = high95,
             color = model, fill = model)) +
  geom_line() + 
  geom_ribbon(alpha = 0.2) +
  theme_classic() +
  scale_x_date(date_breaks = "months" , date_labels = "%b") +
  # ylim(0, 1500) +
  labs(y = "Cumulative hospitalized",
       x = "",
       title = "Prior versus Posterior",
       subtitle = "Difference in models indicates the influence of the data") +
  # ylim(0, 25000) +
  NULL
```


