---
title: "COVID-19 modeling in South Dakota"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: github_document
---
# Authors
*Jeff Wesner, Ph.D.*^1^, *Dan Van Peursem, Ph.D.*^2^, *Jose Flores, Ph.D.*^2,3^, *Yuhlong Lio, Ph.D.*^2^

University of South Dakota

^1^Department of Biology, ^2^Department of Mathematical Sciences, ^3^Department of Computer Science

<Jeff.Wesner@usd.edu>

# Purpose
To predict hospital bed needs, ICU needs, and ventilator needs in South Dakota due to COVID-19. 

# Updates

# Alternative Approach
## Directly modeling hospital bed use

Our previous approach attempted to estimate hospital bed use from an initial SIR (https://github.com/jswesner/covid_sd/blob/master/code/alternative_analysis.pdf). However, there is now enough data to directly model the curve of cumulative COVID related hospitalizations. To do that, we first used the Weibull equation to fit a Bayesian non-linear curve to cumulative hospitalization data (first graph below). We then converted that cumulative prediction to active hospitalizations by varying the days in the hospital to fit the data on active hospitalizations.   
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(viridis)
library(deSolve)
library(scales)
library(lubridate)
library(viridis)
library(readr)
library(readxl)
library(ggrepel)
library(cowplot)
library(viridis)
library(brms)
library(kableExtra)
```


```{r echo=FALSE, fig.height=2, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot() +
  annotate("text", x = 0.5, y = 0.7,
                   label = "(y_cumulative[i]) %~% italic(N)(mu[i], sigma)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.52,
           label = "italic(mu[i]) == frac(italic(max)*date[i]^omega, theta^omega + date[i]^omega)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.35,
           label = "italic(max) %~% italic(beta)(2.76, 549)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.27,
           label = "italic(theta) %~% italic(Gamma)(4, 0.04)", parse = T, family = "serif") +
   annotate("text", x = 0.5, y = 0.19,
           label = "italic(omega) %~% italic(N)(4, 0.4)", parse = T, family = "serif") +
  annotate("text", x = 0.5, y = 0.11,
           label = "sigma %~% italic(HalfCauchy)(0,1)", parse = T, family = "serif") +
  coord_cartesian(ylim = c(0.11, .8)) +
  labs(title = "Weibull non-linear regression:") +
  theme_void() +
  theme(text = element_text(family = "serif"))
```

Prior values for are loosely derived from from New York City's hospitalization curve in which ~ 0.4% of the population was hospitalized. We also bounded the proportion of hospitalizations at zero to prevent predictions of negative hospitalizations.

## Weibull results

Results from the model suggest that South Dakota either has or shortly will reach its peak in COVID-related hospitalizations at ~80 people actively hospitalized.

```{r include=FALSE, cache = F}
sd_pop <- 903027 #from here: https://worldpopulationreview.com/states/south-dakota-population/

sd_covid_data <- read.csv(here::here("data/data_kelo.csv")) %>%
  clean_names() %>%
  mutate(date_num = as.numeric(mdy(date)) - 18329,
         date = mdy(date),
         source = "SD_DOH via Keloland news",
         incidence = positive_cases - lag(positive_cases),
         cum_hosp = parse_number(as.character(hospitalized)),
         incidence1 = incidence + 1,
         cum_prop_hosp = cum_hosp/sd_pop)  ##adds 1 to all incidence to allow for zeros in log-linear regression


all_hosp <- sd_covid_data %>% select(date, hospitalized_currently) %>% arrange(date)



#hosp_beds
sd_hospital_beds <- read_excel(here::here("data/sd_hospital_beds.xlsx")) %>% 
  clean_names()

all_beds <- sum(sd_hospital_beds$total_beds)

#nyc data from here: https://www1.nyc.gov/site/doh/covid/covid-19-data.page
nyc_hosp <- nyc_hosp <- read_csv("data/nyc_hosp.csv") %>% clean_names() %>% 
  mutate(cum_hosp = cumsum(hospitalizations),
         nyc_pop = 8900000,
         date = mdy(date_of_interest),
         date_num = as.numeric(date) - 18323,
         cum_prop_hosp = cum_hosp/nyc_pop) %>% 
  mutate(cum_hosp_rate = cum_hosp/max(cum_hosp),
         date_rate = date_num/max(date_num))

max_prop_hosp <- max(nyc_hosp$cum_prop_hosp)
sd_prior_max = sd_pop*max_prop_hosp
sd_prior_maxsd = 2000 


# fit weibull model 

beta_parms <- function(mean, sd) {
  out <- tibble(alpha = (mean^2 - mean^3 - mean*sd^2)/sd^2,
                beta = (mean - 2*mean^2 + mean^3 - sd^2 + mean*sd^2)/sd^2)
  
  plot <- plot(density(rbeta(1000, out[[1]], out[[2]])))
  return(out)
  return(plot)
}

gamma_parms <- function(mean, sd) {
  out <- tibble(shape = mean^2/sd^2,
                rate = mean/sd^2,
                scale = 1/(mean/sd^2))
  return(out)
  return(plot)
}

beta_parms(mean = 0.005, sd = 0.003)
gamma_parms(mean = 150, sd = 50)


# fit_weib <- brm(bf(cum_prop_hosp|trunc(lb = 0) ~ (ult*date_num^omega)/(theta^omega + date_num^omega),
#                    ult ~ 1, omega ~ 1, theta ~ 1,
#                    nl = TRUE),
#                 data = sd_covid_data,
#                 # family = gaussian(),
#                 family = gaussian(),
#                 prior = c(
#                   prior(beta(2.76, 549), nlpar = "ult"), #original_prior
#                   prior(gamma(4, 0.04), nlpar = "theta"),
#                   prior(normal(4, 0.4), nlpar = "omega")),
#                 iter = 4000,
#                 sample_prior = "yes")

fit_priorday <- readRDS(here::here("code/fit_weib.rds"))
fit_weib <- update(fit_priorday, newdata = sd_covid_data) #re-run the model above with today's data (this shortcut just prevents the need to recompile the model)

# priors come from here:
# ult is the Asymptote: Prior assumes asymptote with mean of 4000 cumulative total people ever hospitalized with wide sd of 2000 - parameterized as beta. That is
# based on NYC's curve, in which a total of 0.45% of their population was ever hospitalized.

# theta is the inflection point - Prior assumes mean of 100 with sd of 30 days - parameterized as gaussian. 
# omega is the slope of inflection point - Prior assumes mean of 50 with sd of 20 - parameterized as gamma.


#predictions up to 200 days after the first hospitalization
newdata = tibble(date_num = 0:200)

predict_cumulative <- predict(fit_weib, newdata = newdata, probs = c(0.025, 0.975, 0.25, 0.75))  %>% 
  as_tibble() %>% 
  clean_names() %>% 
  mutate(date_num = 18329 + newdata$date_num,
         date = as_date(date_num),
         method = "predicted") 



#make predictions of daily hospital needs
days_in_hosp <- 10 #adjusted to visually match actual active hosps

predict_dailyhosp <- predict(fit_weib, newdata = newdata, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_num = 18328 + parse_number(key),
         date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(daily_total = value - lag(value, days_in_hosp)) %>%
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  mutate(daily_total = daily_total*sd_pop) %>%
  group_by(iter, date) %>% 
  filter(!any(daily_total < 0)) %>% 
  group_by(date) %>% 
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high95 = quantile(daily_total, probs = 0.975),
            high50 = quantile(daily_total, probs = 0.75),
            low95 = quantile(daily_total, probs = 0.025),
            low50 = quantile(daily_total, probs = 0.25))
```
```{r echo=FALSE, fig.height=9, fig.width=8, message=FALSE, warning=FALSE, paged.print=FALSE}
# Make plots

cumulative_plot <- predict_cumulative %>% 
  ggplot(aes(x = date, y = estimate*sd_pop)) +
  geom_line() +
  geom_ribbon(aes(ymin = q2_5*sd_pop, ymax = q97_5*sd_pop), alpha = 0.2 , fill = "dodgerblue") +
  geom_ribbon(aes(ymin = q25*sd_pop, ymax = q75*sd_pop), alpha = 0.2 , fill = "dodgerblue")+
  geom_point(data = sd_covid_data, aes(x = date, y = cum_prop_hosp*sd_pop),
             shape = 21, fill = "yellow") +
  theme_classic() +
  scale_x_date(date_breaks = "months" , date_labels = "%b") +
  # ylim(0, 1500) +
  labs(y = "Cumulative hospitalized",
       title = "Weibull model results",
       subtitle = "Dots = data. Shading = 50% and 95% prediction intervals") +
  NULL


active_plot <- predict_dailyhosp %>% 
  ggplot() +
  geom_line(aes(x = date, y = median)) +
  geom_ribbon(aes(x = date, ymin = low95, ymax = high95), alpha = 0.2, fill = "dodgerblue") +
  geom_ribbon(aes(x = date, ymin = low50, ymax = high50), alpha = 0.2, fill = "dodgerblue") +
  geom_point(data = all_hosp, aes(x = date, y = hospitalized_currently),
             shape = 21, fill = "yellow") +
  scale_x_date(date_breaks = "months" , date_labels = "%b") +
  theme_classic() +
  # geom_hline(yintercept = all_beds, color = "dodgerblue") +
  # annotate("text", x = as.Date("2020-04-10"), y = 2960, label = "All Hospital Beds\nin SD") +
  labs(y = "Active hospitalizations",
       title = "Predicting active hospitalizations from Weibull model",
       subtitle = "Dots = data. Shading = 50% and 95% prediction intervals") +
  coord_cartesian(ylim = c(0,100)) +
  NULL


both_plot_weib <- plot_grid(cumulative_plot, active_plot, ncol = 1)

both_plot_weib
```

## Influence of the priors
The plot below compares the prior predictive distribution (i.e. running the model with only the prior information, not the data) to the posterior predictive distribution. The large difference in these predictions indicates that we learned a lot of information from the data and the priors have relatively small influences.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#prior v post

posts <- posterior_samples(fit_weib) %>% as_tibble() %>% mutate(iter = 1:nrow(.))

x = seq(1,200, by = 10)

priors <- posts %>% select(prior_b_ult, prior_b_omega, prior_b_theta, iter) %>% sample_n(1000) %>% 
  expand_grid(x) %>% 
  mutate(pred = (prior_b_ult*x^prior_b_omega)/(prior_b_theta^prior_b_omega + x^prior_b_omega),
         model = 'prior_prediction') %>% 
  select(x, pred, model)

posteriors <- posts %>% select(b_ult_Intercept, b_omega_Intercept, b_theta_Intercept) %>% sample_n(1000) %>% 
  expand_grid(x) %>% 
  mutate(pred = (b_ult_Intercept*x^b_omega_Intercept)/(b_theta_Intercept^b_omega_Intercept + x^b_omega_Intercept),
         model = 'posterior_prediction') %>% 
  select(x, pred, model)


all_posts <- bind_rows(priors, posteriors) %>% 
  as_tibble() %>% 
  mutate(date_num = 18329 + x,
         date = as_date(date_num))

all_posts %>% 
  group_by(date, model) %>% 
  summarize(mean = mean(pred),
            low95 = quantile(pred, probs = 0.025),
            high95 = quantile(pred, probs = 0.975)) %>% 
  ggplot(aes(x = date, y = mean*sd_pop, ymin = low95*sd_pop, ymax = high95*sd_pop,
             color = model, fill = model)) +
  geom_line() + 
  geom_ribbon(alpha = 0.2) +
  theme_classic() +
  scale_x_date(date_breaks = "months" , date_labels = "%b") +
  # ylim(0, 1500) +
  labs(y = "Cumulative hospitalized",
       x = "",
       title = "Prior versus Posterior",
       subtitle = "Difference in models indicates the influence of the data") +
  NULL
```


## Outside of Sioux Falls
Sioux Falls, SD had the earliest large outbreak in South Dakota. To determine how COVID19-related hospitalizations differed in Sioux Falls and outside of Sioux Falls, we ran the model above separately - once for Minnehaha County (where Sioux Falls is located) and once for all other areas of the state. The results are plotted below. They suggested that Minnehaha County is near or past its peak hospitalizations, but that the rest of South Dakota might not peak until mid-late summer or later.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, include = F}
library(tidyverse)
library(here)
library(janitor)
library(viridis)
library(deSolve)
library(scales)
library(lubridate)
library(viridis)
library(readr)
library(readxl)
library(ggrepel)
library(cowplot)
library(viridis)
library(brms)
library(kableExtra)


cum_hosp <- read_csv("data/cum_hosp.csv", na = c("", ".")) %>% mutate(date = mdy(date_hosp))
sd_county_pops <- read_csv("data/sd_county_pops.csv") %>% 
  select(-X1) 
# 
# new_data <- tibble(date_hosp = ymd("2020-05-15"),
#                    group = c("Minnehaha County", "Rest of South Dakota"),
#                    n = NA,
#                    cum_hosp = c(228, 76))

group_pops <- sd_county_pops %>% 
  mutate(group = case_when(county == "Minnehaha" ~ "Minnehaha County",
                           TRUE ~ "Rest of South Dakota")) %>%
  group_by(group) %>% 
  summarize(group_pop = sum(total_pop))


# beta_parms <- function(mean, sd) {
#   out <- tibble(alpha = (mean^2 - mean^3 - mean*sd^2)/sd^2,
#                 beta = (mean - 2*mean^2 + mean^3 - sd^2 + mean*sd^2)/sd^2)
#   
#   plot <- plot(density(rbeta(1000, out[[1]], out[[2]])))
#   return(out)
#   return(plot)
# }

# beta_parms(0.0004, 0.0001)
# fit_weib <- brm(bf(cum_prop_hosp|trunc(lb = 0) ~ (ult*date_num^omega)/(theta^omega + date_num^omega),
#                    ult ~ 1, omega ~ 1, theta ~ 1,
#                    nl = TRUE),
#                 data = sd_covid_data,
#                 # family = gaussian(),
#                 family = gaussian(),
#                 prior = c(
#                   prior(beta(2.76, 549), nlpar = "ult"), #original_prior
#                   prior(gamma(4, 0.04), nlpar = "theta"),
#                   prior(normal(4, 0.4), nlpar = "omega")),
#                 iter = 4000,
#                 sample_prior = "yes")
minn_hosp <- cum_hosp %>% filter(group == "Minnehaha County") %>% rename(cum_prop_hosp = prop_hosp) %>% mutate(date_num = date_num - 18341)
rest_hosp <- cum_hosp %>% filter(group == "Rest of South Dakota") %>% rename(cum_prop_hosp = prop_hosp) %>% mutate(date_num = date_num - 18329)


fit_priorday <- readRDS(here::here("code/fit_weib.rds"))

fit_minn <- update(fit_priorday, newdata = minn_hosp) #re-run the model above with today's data (this shortcut just prevents the need to recompile the model)
# post_minn <- posterior_samples(fit_minn)
# beta_parms(mean = mean(post_minn$b_ult_Intercept), sd = 2*sd(post_minn$b_ult_Intercept)) #prior for rest of sd

fit_rest <- update(fit_priorday, newdata = rest_hosp, prior = c(
                  prior(beta(679, 476819), ub = 1, lb = 0, nlpar = "ult"), # use prior from Minnehaha for the max but with 2x wider sd
                  prior(gamma(4, 0.04), lb = 0, nlpar = "theta"),
                  prior(normal(4, 0.4), nlpar = "omega")))

# priors come from here:
# ult is the Asymptote: Prior assumes asymptote with mean of 4000 cumulative total people ever hospitalized with wide sd of 2000 - parameterized as beta. That is
# based on NYC's curve, in which a total of 0.45% of their population was ever hospitalized.

# theta is the inflection point - Prior assumes mean of 100 with sd of 30 days - parameterized as gaussian. 
# omega is the slope of inflection point - Prior assumes mean of 50 with sd of 20 - parameterized as gamma.
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#predictions up to 200 days after the first hospitalization
newdata = tibble(date_num = 0:300)

predict_cumulative_minn <- predict(fit_minn, newdata = newdata, probs = c(0.025, 0.975, 0.25, 0.75))  %>% 
  as_tibble() %>% 
  clean_names() %>% 
  mutate(date_num = newdata$date_num,
         date = as_date(date_num + 18341),
         method = "predicted",
         group = "Minnehaha County") 

predict_cumulative_rest <- predict(fit_rest, newdata = newdata, probs = c(0.025, 0.975, 0.25, 0.75))  %>% 
  as_tibble() %>% 
  clean_names() %>% 
  mutate(date_num = newdata$date_num,
         date = as_date(date_num + 18329),
         method = "predicted",
         group = "Rest of South Dakota") 


all_cumulative <- bind_rows(predict_cumulative_minn, predict_cumulative_rest)

#make predictions of daily hospital needs
days_in_hosp <- 10 #adjusted to visually match actual active hosps
#
predict_dailyhosp_minn <- predict(fit_minn, newdata = newdata, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_num = 18341 + parse_number(key),
         date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(daily_total = value - lag(value, days_in_hosp)) %>%
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  mutate(daily_total = daily_total) %>%
  group_by(iter, date) %>%
  filter(!any(daily_total < 0)) %>%
  group_by(date) %>%
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high95 = quantile(daily_total, probs = 0.975),
            high50 = quantile(daily_total, probs = 0.75),
            low95 = quantile(daily_total, probs = 0.025),
            low50 = quantile(daily_total, probs = 0.25))



predict_dailyhosp_rest <- predict(fit_rest, newdata = newdata, summary = F)  %>%
  as_tibble() %>%
  mutate(iter = 1:nrow(.)) %>%
  gather(key, value, -iter) %>%
  mutate(date_num = 18328 + parse_number(key),
         date = as_date(date_num)) %>%
  arrange(iter, date_num) %>%
  group_by(iter) %>%
  mutate(daily_total = value - lag(value, days_in_hosp)) %>%
  group_by(date_num) %>%
  drop_na(daily_total) %>%
  mutate(daily_total = daily_total) %>%
  group_by(iter, date) %>%
  filter(!any(daily_total < 0)) %>%
  group_by(date) %>%
  summarize(median = median(daily_total),
            mean = mean(daily_total),
            sd = sd(daily_total),
            high95 = quantile(daily_total, probs = 0.975),
            high50 = quantile(daily_total, probs = 0.75),
            low95 = quantile(daily_total, probs = 0.025),
            low50 = quantile(daily_total, probs = 0.25))

all_daily <- bind_rows(predict_dailyhosp_minn %>% mutate(group = "Minnehaha County"),
                       predict_dailyhosp_rest %>% mutate(group = "Rest of South Dakota"))



# Make plots
cumulative_plot <- all_cumulative %>% 
  left_join(group_pops) %>% 
  ggplot(aes(x = date, y = estimate*group_pop)) +
  geom_line(aes(color = group)) +
  geom_ribbon(aes(ymin = q2_5*group_pop, ymax = q97_5*group_pop, fill = group), alpha = 0.2 ) +
  geom_ribbon(aes(ymin = q25*group_pop, ymax = q75*group_pop, fill = group), alpha = 0.2 ) +
  geom_point(data = cum_hosp, aes(x = date, y = cum_hosp, fill = group),
             shape = 21) +
  # facet_wrap(~ group, scales = "free") +
  theme_classic() +
  scale_x_date() +
  # scale_x_date(date_breaks = "months" , date_labels = "%b") +
  # ylim(0, 1500) +
  labs(y = "Cumulative ever hospitalized",
       title = "Weibull model results",
       subtitle = "Dots = data. Shading = 50% and 95% prediction intervals") +
  NULL

# cumulative_plot


active_plot <- all_daily %>% 
  left_join(group_pops) %>% 
  ggplot() +
  geom_line(aes(x = date, y = median*group_pop, color = group)) +
  geom_ribbon(aes(x = date, ymin = low95*group_pop, ymax = high95*group_pop, fill = group), alpha = 0.2) +
  geom_ribbon(aes(x = date, ymin = low50*group_pop, ymax = high50*group_pop, fill = group), alpha = 0.2) +
  # geom_point(data = all_hosp, aes(x = date, y = hospitalized_currently),
  #            shape = 21, fill = "yellow") +
  # scale_x_date(date_breaks = "months" , date_labels = "%b") +
  theme_classic() +
  # geom_hline(yintercept = all_beds, color = "dodgerblue") +
  # annotate("text", x = as.Date("2020-04-10"), y = 2960, label = "All Hospital Beds\nin SD") +
  labs(y = "Active hospitalizations",
       title = "Predicting active hospitalizations from Weibull model",
       subtitle = "Shading = 50% and 95% prediction intervals") +
  # coord_cartesian(ylim = c(0,100)) +
  NULL
# active_plot

both_plot_group <- plot_grid(cumulative_plot, active_plot, ncol = 1, align = "v")

both_plot_group
ggsave(both_plot_group, file = "plots/both_plot_group.jpg", width = 7, height = 7)
```

